Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-06 VegOzone.tex
Restructured sec 3 - texts has not been fully revised yet

%%
\section{Data acquisition of ozone and key environmental variables}
\label{sec:data}

@@

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019. For further analysis (Section~\ref{sec:stats}) and modeling study with the $\mathrm{DO_3SE}$ model (Section~\ref{sec:do3se}), we reconstructed the missing data. The used gap filling method will be described in detail in a companion paper to be submitted.

Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance ($Q_0$) displayed in Fig.~\ref{fig:data_svanvik_2018_2019}d was higher in May and July 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

@@

\section{Statistical analysis of key environmental variables}
\label{sec:stats}
To improve our understanding and determine the risk of vegetation from high \chem{[O_3]} in subarctic biomes in a changing climate, we first assess the climate conditions and evaluate the growing seasons 2018/19 concerning anomalies from multi-annual means (climatologies). We derive climatologies for temperature, precipitation, global irradiance, and ozone for Svanvik. The ozone measurements will be put into a climatological as well as regional context.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth which is shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc) \citep{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

Winter and spring (November--April, except for March) are relatively dry ($\sum_m \mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_m \mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a highest monthly averaged accumulated precipitation $\sum_m \mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large interannual variability.

A climatology derived for global irradiance $Q_0$ is shown in Appendix Fig.~\ref{fig:global_rad_clim}a) for both, climatological average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$ irradiance observed at Svanvik. In agreement with the geometrically derived sunshine duration shown in Supplement Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

@@

Finally, we shall have a look at the climatologies derived from observational ozone data and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). To represent the sites, the nearest neighboring grid point of the \chem{[O_3]} reanalysis products is chosen. The global products do not reproduce observed ground-level \chem{[O_3]} well. This is important for European wide and global risk assessment that rely on these data. The MACC reanalysis (Fig.~\ref{fig:ozone_climatologies}a) is in general too low ($(9\pm 7)\,\unit{ppb}$) and displays an erroneous seasonal cycle. Ozone concentrations peak already in March followed by the second peak in July. Also CAMSRA matches the observed ozone climatology poorly (Fig.~\ref{fig:ozone_climatologies}b). Despite displaying only one ozone peak, it still does not reproduce the seasonality in northern Fennoscandia well. The CAMSRA spring peak lags behind observations by $1\,\unit{month}$ and is $5\,\unit{ppb}$ too low, whereas the minimum occurs in January compared to August/September. The annual amplitude ($(26\pm 1)\,\unit{ppb}$) is larger than in the climatology derived from observations ($19\,\unit{ppb}$). The modeled \chem{[O_3]} at Svanvik are slightly higher than at Pallas which might be due to the height at which surface ozone has been evaluated in the model. In contrast, the ensemble mean of the CAMSRAQ reproduces the seasonal cycle in northern Fennoscandia quite well (Fig.~\ref{fig:ozone_climatologies}c), but shows on average a slight low-bias ($(2.8\pm 0.5)\,\unit{ppb}$).

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatologies_portrait}
  \includegraphics[width=8.3cm]{fig07}
  \caption{Comparison of climatologies derived from observational data (gray band) with three ozone reanalysis products: (a) MACC reanalysis; (b) CAMSRA; (c) CAMSRAQ reanalysis ensemble mean. The sites, represented by the nearest neighboring grid point of the \chem{[O_3]} reanalysis products, are displayed in different markers and colors. The global products (a), (b) do not reproduce observed ground-level ozone well. The MACC reanalysis (a) is in general too low and displays an erroneous seasonal cycle. The CAMSRA (b) matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of CAMSRAQ reanalysis ensemble reproduces the seasonal cycle in northern Fennoscandia, though the modeled \chem{[O_3]}s are slightly too low.}
  \label{fig:ozone_climatologies}
\end{figure}

There are many probable explanations for the discovered differences in the climatologies derived from the three reanalyses. The model enhancements from MACC reanalysis to CAMSRA have been reported and discussed by \citet{ACP:Inness2019} on a global level. Amongst others, assimilation of ozone profiles from satellite retrieval and an upgraded ozone chemistry scheme has lead to the enhanced performance of the latter, but a considerable bias still remains \citep{GMD:Huijnen2020}. In particular, \citet{ACPD:Barten2020} reported a pronounced bias for CAMSRA in the high Arctic (e.g., Summit, Greenland) and attribute this bias to dry deposition to the ocean. The higher spatial resolution of the regional air quality models used in the CAMSRAQ ensemble compared to the global reanalyses captures small scale depletion and peak episodes in \chem{[O_3]} better and enhances the capability of the models concerning daily and seasonal cycles in ozone. The latter is, of course, also affected by upgrades in the OpenIFS used as meteorological forcing. However, we can assume that the main driver for the strikingly different results is the assimilated observational ozone data. Satellite retrievals resolve \chem{[O_3]} at the surface rather poorly and hence do not constrain the global models well enough \citep{ACP:Andersson2017}.

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
Using these established climatologies, we look at the anomalies in 2018/19 and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent in occurrence with climate change, this will help improving our understanding of the role which these climate variables have on determining the sensitivity of subarctic vegetation to \chem{[O3]}.

@@

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and estimate the ozone concentration anomalies for 2018/19. In Fig.~\ref{fig:anomalies_svanvik}d), the resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month. Negative deviations from climatology are shown as a negative percentage. Note that missing bars for certain months are both due to data availability (see Fig~\ref{fig:data_svanvik_2018_2019} for details concerning Svanvik) and anomalies not meeting the criteria of $\pm 1\sigma$ deviation from the climatological mean. In April/May 2018, ozone was significantly enhanced, while 2019 was generally an average year. In both years, ozone concentrations were significantly below average in November. July 2018 has not been corrected for the missing data due to data acquisition problems. The corrected value is, however, indicated by the star (refer to Section~\ref{subsec:ozone_reco} for details of the used method for gap-filling).

>>>>

%%
\section{Data acquisition}
\label{sec:data}

@@

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019. For further analysis (Section~\ref{sec:stats}) and modeling study with the $\mathrm{DO_3SE}$ model (Section~\ref{sec:do3se}), we reconstructed the missing data. The used gap-filling method will be described in detail in a companion paper to be submitted.

Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance ($Q_0$) displayed in Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in May and July 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions and multi-annual mean (referred to as climatology in thhe following) annual cycle of gound-level ozone concentrations at Svanhovd. To this end, we derive climatologies for temperature, precipitation, global irradiance, and ozone. We evaluate statistical significance of divergances from the norm in these variables for the growing seasons 2018/19 (referred to as anomalies). We perform a bias correction on the Svanvik ozone climatology with respect to long-term series of \chem{[O_3]} at the sites in northern Finland and Sweden.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth which is shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from the atmospheric monitoring in Svanvik (1992--2012). Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc) \citep{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

@@

The average accumulated monthly precipitation $\sum_d\mathrm{Precip}$ indicates that winter and spring (November--April, except for March) are relatively dry ($\sum_d\mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_d\mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a $\sum\mathrm{Precip}_d = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation ($\sum_m\mathrm{Precip}$) given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large interannual variability.

The derived climatology for global irradiance ($Q_0$) at Svanvik is shown in Appendix Fig.~\ref{fig:global_rad_clim}a). Shown are the average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$. In agreement with the geometrically derived sunshine duration displayed in Supplement Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
Using these established climatologies, we look at the anomalies in 2018/19 and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to improve our understanding of a possible future increase in vulnerability of subarctic vegetation to ozone.

@@

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and estimate the ozone concentration anomalies for 2018/19. In Fig.~\ref{fig:anomalies_svanvik}d), the resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month. Negative deviations from climatology are shown as a negative percentage. Note that missing bars for certain months are both due to data availability (see Fig~\ref{fig:data_svanvik_2018_2019} for details concerning Svanvik) and anomalies not meeting the criteria of $\pm 1\sigma$ deviation from the climatological mean. In April/May 2018, ozone was significantly enhanced, while 2019 was generally an average year. In both years, ozone concentrations were significantly below average in November. July 2018 represented by the color bar has not been corrected for the missing data due to data acquisition problems. The corrected value is indicated by the star. A detailed describtion of the gap-filling based on a Reynolds's decomposition method will be published seperately.
